---
title: "1.3_Basic EDA in R"
format: html
editor: visual
---

### Introduction to the Superstore Dataset

#### Overview of the Superstore Dataset

The Superstore dataset is a widely-used fictional dataset created by Tableau, designed to simulate sales data from a retail store. The dataset contains a comprehensive set of records capturing various aspects of sales transactions, customer information, and product details.

#### Fields/Columns in the Dataset

The Superstore dataset consists of several fields/columns, each representing a different aspect of the sales transactions and related information. Here is an overview of the key fields/columns present in the dataset:

1.  **Row ID**: Unique identifier for each row in the dataset.
2.  **Order ID**: Unique identifier for each sales order.
3.  **Order Date**: Date when the order was placed.
4.  **Ship Date**: Date when the order was shipped.
5.  **Ship Mode**: Shipping method used for the order (e.g., Standard, Express).
6.  **Customer ID**: Unique identifier for each customer.
7.  **Customer Name**: Name of the customer.
8.  **Segment**: Market segment to which the customer belongs (e.g., Consumer, Corporate, Home Office).
9.  **Country**: Country where the order was placed.
10. **City**: City where the order was placed.
11. **State**: State or province where the order was placed.
12. **Postal Code**: Postal code of the location where the order was placed.
13. **Region**: Geographic region where the order was placed.
14. **Product ID**: Unique identifier for each product.
15. **Category**: Category to which the product belongs (e.g., Furniture, Office Supplies, Technology).
16. **Sub-Category**: Sub-category of the product (e.g., Chairs, Phones, Accessories).
17. **Product Name**: Name of the product.
18. **Sales**: Total sales amount for the order.
19. **Quantity**: Number of units sold.
20. **Discount**: Discount applied to the order.
21. **Profit**: Profit generated from the order.

Understanding the meaning and significance of each field/column in the Superstore dataset is essential for performing effective exploratory data analysis and deriving meaningful insights from the data.

### Loading and Inspecting the Data

#### Importing the Superstore Dataset into R

To start our exploratory data analysis (EDA) with the Superstore dataset, we need to import it into R. Here's a step-by-step guide to importing the dataset:

1.  **Load Necessary Packages**: Before importing the dataset, ensure you have the necessary packages installed. The `readr` package is commonly used for reading data files in R. If you haven't installed it yet, you can do so using `install.packages("readr")`.

```{r}
# Load the 'readr' package
library(readr)
```

2.  **Specify File Path**: Define the file path of the Superstore dataset. Replace `"path_to_superstore_dataset.csv"` with the actual file path where the dataset is located on your system.

```{r}
# Specify the file path of the Superstore dataset
file_path <- "data\\superstore.csv"
```

3.  **Import Dataset**: Use the `read_csv()` function from the `readr` package to import the dataset into R. Assign the imported dataset to a variable, such as `superstore_data`.

```{r}
# Import the dataset into R
superstore_data <- read_csv(file_path)
```

4.  **Check Class of Imported Object**: Verify the class of the imported object using the `class()` function. This confirms that the dataset has been successfully imported.

```{r}
# Check the class of the imported object
class(superstore_data)
```

#### Basic Data Inspection Techniques

Once the dataset is imported, we can employ various techniques to inspect its structure and content:

1.  **head() Function**: The `head()` function displays the first few rows of the dataset, providing a glimpse into its contents. This is useful for quickly assessing the variables and their values.

```{r}
# Display the first few rows of the dataset
head(superstore_data)
```

2.  **tail() Function**: Similar to `head()`, the `tail()` function displays the last few rows of the dataset. This can be helpful for examining the end of the data.

```{r}
# Display the last few rows of the dataset
tail(superstore_data)
```

3.  **summary() Function**: The `summary()` function generates summary statistics for each numerical variable in the dataset. It calculates metrics such as mean, median, minimum, maximum, and quartiles, providing insights into the distribution of numeric data.

```{r}
# Generate summary statistics for numerical variables
summary(superstore_data)
```

4.  **str() Function**: The `str()` function offers a concise summary of the dataset's structure. It displays the data types and dimensions of each variable, along with a preview of the data.

```{r}
# Display the structure of the dataset
str(superstore_data)
```

By employing these basic data inspection techniques, we can gain a comprehensive understanding of the Superstore dataset, enabling us to proceed with further analysis and exploration.

If `read_csv()` reads dates as character columns instead of Date, we can specify the column types explicitly using the `col_types` argument. Here's how you can fix it:

```{r}
library(readr)

# Read the CSV file with explicit column types
superstore_data <- read_csv(file_path)
```

In the above code:

-   `col_types` is used to specify the types of each column in the dataset.
-   `OrderDate = col_date(format = "%Y-%m-%d")` specifies that the `OrderDate` column should be read as a date column, with the specified date format.
-   `.default = col_character()` specifies that all other columns should be read as character columns by default.

Adjust the column names and formats according to your actual dataset structure. This ensures that the dates are correctly interpreted as Date objects in R.

We can also update column datatypes after we open the dataset

```{r}
# Convert Order Date and Ship Date to Date objects
superstore_data$`Order Date` <- as.Date(superstore_data$`Order Date`, format = "%m/%d/%Y")
superstore_data$`Ship Date` <- as.Date(superstore_data$`Ship Date`, format = "%m/%d/%Y")

# Convert Segment and Country to factors
superstore_data$Segment <- as.factor(superstore_data$Segment)
superstore_data$Country <- as.factor(superstore_data$Country)
```
