---
title: "Basic Flow of Control in R"
format: html
editor: visual
---

#### Repetition in R (Loops)

Loops are fundamental constructs in programming that allow you to execute a block of code repeatedly. In R, the primary looping constructs are the `for` loop, the `while` loop, and the `repeat` loop. This section will guide you through the syntax and usage of each, with examples to illustrate how they can be used in data analysis tasks.

#### The `for` Loop

The `for` loop is used to iterate over a sequence or a vector and execute a block of code for each element.

##### Syntax:

```{r, eval=FALSE}
for (variable in sequence) {
  # Code to execute for each element of the sequence
}
```

##### Example:

Let's iterate over a vector of numbers and print each number squared.

```{r}
# Create a vector of numbers
numbers <- 1:5

# Use a for loop to iterate and print squares
for (num in numbers) {
  print(num^2)
}
```

What are the correct values for the \*s in the code below so that it will iterate over all the items of x array?

```{r, eval=FALSE}

x <- 50:60
for (i in *:*)
{
  print(x[i])
}
```

In this example, the loop runs five times, squaring each number from 1 to 5.

#### The `while` Loop

The `while` loop continues to execute as long as a specified condition is true. It's useful when the number of iterations is not known beforehand.

##### Syntax:

```{r, eval=FALSE}
while (condition) {
  # Code to execute as long as the condition is true
}
```

##### Example:

We will use a `while` loop to simulate a countdown.

```{r}
# Set the starting point of the countdown
countdown <- 5

# Begin the while loop
while (countdown > 0) {
  print(paste("Counting down:", countdown))
  countdown <- countdown - 1
}

# When the loop ends
print("Lift off!")
```

This loop will run until `countdown` is no longer greater than 0.

#### The `repeat` Loop

The `repeat` loop executes an indefinite number of times until explicitly told to stop with a `break` statement. It's the most basic form of loop and is less commonly used than `for` and `while` loops due to its potential to create infinite loops if not handled carefully.

##### Syntax:

```{r, eval=FALSE}
repeat {
  # Code to execute indefinitely
  if (stop_condition) {
    break
  }
}
```

##### Example:

Here's a `repeat` loop that finds the first number divisible by 13 greater than 100.

```{r}
# Initialize the number
number <- 101

# Start the repeat loop
repeat {
  # Check if the number is divisible by 13
  if (number %% 13 == 0) {
    print(paste("Found a number divisible by 13:", number))
    break
  }
  # Increment the number
  number <- number + 1
}
```

The loop runs indefinitely, incrementing `number` until it finds one that satisfies the condition.

#### Nested Loops

Loops can be nested inside other loops, which is useful for iterating over multiple dimensions, such as rows and columns in a matrix.

##### Example:

```{r}
# Create a 3x3 matrix
my_matrix <- matrix(1:9, nrow = 3)

# Iterate over each element of the matrix
for (row in 1:nrow(my_matrix)) {
  for (col in 1:ncol(my_matrix)) {
    # Multiply each element by 2
    my_matrix[row, col] <- my_matrix[row, col] * 2
  }
}
print(my_matrix)
```

In this nested loop example, every element of `my_matrix` is doubled.

### Conditional Branching in R

#### Introduction to Conditional Branching

Conditional branching in R allows you to control the flow of execution based on conditions. It's a fundamental concept in programming that lets you execute certain sections of code while skipping others, depending on whether specified conditions are `TRUE` or `FALSE`. This section will explain how to use conditional statements in R, such as `if`, `else`, and `else if`, along with practical examples.

#### The `if` Statement

The `if` statement is the simplest form of conditional branching. It evaluates a condition and, if that condition is `TRUE`, it executes a block of code.

##### Syntax:

```{r, eval=FALSE}
if (condition) {
  # Code to execute if the condition is TRUE
}
```

##### Example:

Here, we check if a number is positive.

```{r}
number <- 5

if (number > 0) {
  print("The number is positive.")
}
```

#### The `else` Statement

The `else` statement is used together with an `if` statement to execute code when the `if` condition is not `TRUE`.

##### Syntax:

```{r, eval=FALSE}
if (condition) {
  # Code to execute if the condition is TRUE
} else {
  # Code to execute if the condition is FALSE
}
```

##### Example:

Expanding on the previous example, we also handle the case when the number is not positive.

```{r}
number <- -3

if (number > 0) {
  print("The number is positive.")
} else {
  print("The number is not positive.")
}
```

#### The `else if` Statement

The `else if` statement allows you to check multiple conditions. If the condition for `if` is `FALSE`, it checks the condition for `else if`, and so on.

##### Syntax:

```{r, eval=FALSE}
if (condition1) {
  # Code to execute if condition1 is TRUE
} else if (condition2) {
  # Code to execute if condition1 is FALSE and condition2 is TRUE
} else {
  # Code to execute if none of the above conditions are TRUE
}
```

##### Example:

Here, we classify a number as positive, negative, or zero.

```{r}
number <- 0

if (number > 0) {
  print("The number is positive.")
} else if (number < 0) {
  print("The number is negative.")
} else {
  print("The number is zero.")
}
```

#### The `switch` Statement

The `switch` statement is a multi-way branch used when you want to compare the same variable (or expression) with many different possible values and execute different pieces of code for each value.

##### Syntax:

```{r, eval=FALSE}
switch(EXPR, 
       "case1" = {# Code for case 1},
       "case2" = {# Code for case 2},
       # ...,
       {# Default case code}
)
```

##### Example:

The following example shows how `switch` can be used to print different messages based on the value of a character variable.

```{r}
day <- "Tue"

switch(day,
       "Mon" = {print("Start of the work week")},
       "Tue" = {print("Second day of the work week")},
       "Wed" = {print("Middle of the work week")},
       "Thu" = {print("Approaching the end")},
       "Fri" = {print("Last day of the work week")},
       {print("It must be the weekend")}
)
```
